---
output: github_document
title: Forecasting Metadata
---

#Example using output from a FLARE forecast

```{r setup, message=FALSE}
library(EML)
library(tidyverse)
library(uuid)
library(ncdf4)
library(lubridate)
emld::eml_version("eml-2.2.0")
```

First, set the forecast identifiers

ForecastProject_id represents the launch of an automated, iterative forecast.  It is created
each time a human modifies the forecast code.  It can be a DOI because this is the level that
we envision citations occuring.

Forecast_id represents each forecast cycle within a ForecastProject_id

For example, if you have a forecast code base on GitHub and launch a forecast from that code that
runs daily for 365 days, then there will be one ForecastProject_id and 365 Forecast_ids. A paper 
analyzing the forecasts would cite the ForecastProject_id.

```{r}

ForecastProject_id <- 30405043 #Some ID that applies to a bunch of forecasts
Forecast_id <- uuid::UUIDgenerate() #The ID that applies to the specific forecast in the cycel
```

# Scenario 1: Oxygen System turned on

## Generating the forecast

First, get the output as we are currently saving it

```{r}
nc <- nc_open("IW2_1_SSSon_H_2019_05_27_2019_06_03_F_16_2132020_21_32.nc")

#Dimnesions
depth <- ncvar_get(nc, "z")
time <- ncvar_get(nc,'time')
local_tzone <- ncatt_get(nc, 0)$time_zone_of_simulation
time <- as.POSIXct(time, origin = '1970-01-01 00:00.00 UTC', tz = local_tzone)
time <-  strftime(time , "%Y-%m-%dT%H:%M:%S%z")

#States
temperature <- ncvar_get(nc, "temp")
oxygen <- ncvar_get(nc, "OXY_oxy")
zone1temp <- ncvar_get(nc, "zone1temp")
zone2temp <- ncvar_get(nc, "zone2temp")
sw_factor <- ncvar_get(nc, "sw_factor")
inflow_factor <- ncvar_get(nc, "inflow_factor")

#Forecast timings
forecasted <- ncvar_get(nc, "forecasted")
data_assimilation <- if_else(forecasted == 0, 1, 0)
#This doesn't work but we need to update our writing script to just have the date
forecast_issue_time <- ncatt_get(nc, varid=0, attname = "history") 

forecast_issue_time <- "2020-02-13T21:32:52-05:00"

nc_close(nc)
```

## Saving to a standardized output format (Option 1)

Second, build a data frame for all variables with a depth dimension

```{r}

n_depths <- length(depth)
state_names <- c("temperature", "oxygen")
states <- list(temperature, oxygen)
n_states <- length(state_names)
df_combined <- list()

for(k in 1:n_states){
for(i in 1:n_depths){
    df <- as_tibble(states[[k]][, ,i])
    names(df) <- as.character(seq(1, ncol(states[[k]][, ,i])))
    df <- cbind(time, df, data_assimilation)
    df <- df %>% 
      pivot_longer(cols = -c(time,data_assimilation), 
                   names_to = "ensemble", 
                   values_to = state_names[k]) %>% 
      mutate(ensemble = as.integer(ensemble)) %>% 
      mutate(depth = depth[i])
    if(i == 1){
    running_df <- df
    }else{
      running_df <- rbind(running_df, df)
    }
}
      df_combined[[k]] <- running_df
}

df_combined <- right_join(df_combined[[1]], df_combined[[2]], 
                          by = c("time", "ensemble", "depth", "data_assimilation")) %>% 
    mutate(forecast_issue_time = forecast_issue_time,
         ForecastProject_id = ForecastProject_id,
         Forecast_id = Forecast_id,
         scenario = "oxygen_on") %>% 
  select(time, depth, scenario, ensemble, temperature, 
         oxygen, forecast_issue_time, 
         data_assimilation, Forecast_id, ForecastProject_id) 

df_combined_scenario_1 <- df_combined

```

Second, build a data frame for all variables without a depth dimension (parameters).

```{r}
df <- as_tibble(zone1temp[ , ])
names(df) <- as.character(seq(1, ncol(zone1temp)))
df <- cbind(time, data_assimilation, df)
df_combined_p1 <- df %>% 
      pivot_longer(cols = -c("time", "data_assimilation"), 
                   names_to = "ensemble", 
                   values_to = "zone1temp") %>% 
      mutate(ensemble = as.integer(ensemble),
             scenario = "oxygen_on") 

df <- as_tibble(zone2temp[ , ])
names(df) <- as.character(seq(1, ncol(zone2temp)))
df <- cbind(time, data_assimilation, df)
df_combined_p2 <- df %>% 
      pivot_longer(cols = -c("time", "data_assimilation"), 
                   names_to = "ensemble", 
                   values_to = "zone2temp") %>% 
      mutate(ensemble = as.integer(ensemble),
             scenario = "oxygen_on") 

df <- as_tibble(sw_factor[ , ])
names(df) <- as.character(seq(1, ncol(sw_factor)))
df <- cbind(time, data_assimilation, df)
df_combined_p3 <- df %>% 
      pivot_longer(cols = -c("time", "data_assimilation"), 
                   names_to = "ensemble", 
                   values_to = "sw_factor") %>% 
      mutate(ensemble = as.integer(ensemble),
             scenario = "oxygen_on") 

df <- as_tibble(inflow_factor[ , ])
names(df) <- as.character(seq(1, ncol(inflow_factor)))
df <- cbind(time, data_assimilation, df)
df_combined_p4 <- df %>% 
      pivot_longer(cols = -c("time", "data_assimilation"), 
                   names_to = "ensemble", 
                   values_to = "inflow_factor") %>% 
      mutate(ensemble = as.integer(ensemble),
             scenario = "oxygen_on") 

df_parameters_1 <- right_join(df_combined_p1, 
                              df_combined_p2, 
                              by = c("time",
                                     "data_assimilation",
                                     "ensemble",
                                     "scenario")) %>% 
                  right_join(df_combined_p3, 
                              by = c("time",
                                     "data_assimilation",
                                     "ensemble",
                                     "scenario")) %>%
                  right_join(df_combined_p4, 
                              by = c("time",
                                     "data_assimilation",
                                     "ensemble",
                                     "scenario")) %>% 
  mutate(forecast_issue_time = forecast_issue_time,
         Forecast_id = Forecast_id,
         ForecastProject_id = ForecastProject_id) %>% 
  select(time, scenario, ensemble, forecast_issue_time, 
         data_assimilation, Forecast_id, ForecastProject_id, 
         zone1temp, zone2temp, sw_factor, inflow_factor) 
```

# Scenario 2: Oxygen System turned off

## Generating the forecast

First, get the output as we are currently saving it

```{r}
nc <- nc_open("IW2_1_SSSoff_H_2019_05_27_2019_06_03_F_16_2132020_21_19.nc")

#Dimnesions
depth <- ncvar_get(nc, "z")
time <- ncvar_get(nc,'time')
local_tzone <- ncatt_get(nc, 0)$time_zone_of_simulation
time <- as.POSIXct(time, origin = '1970-01-01 00:00.00 UTC', tz = local_tzone)
time <-  strftime(time , "%Y-%m-%dT%H:%M:%S%z")

#States
temperature <- ncvar_get(nc, "temp")
oxygen <- ncvar_get(nc, "OXY_oxy")
zone1temp <- ncvar_get(nc, "zone1temp")
zone2temp <- ncvar_get(nc, "zone2temp")
sw_factor <- ncvar_get(nc, "sw_factor")
inflow_factor <- ncvar_get(nc, "inflow_factor")

#Forecast timings
forecasted <- ncvar_get(nc, "forecasted")
data_assimilation <- if_else(forecasted == 0, 1, 0)
#This doesn't work but we need to update our writing script to just have the date
forecast_issue_time <- ncatt_get(nc, varid=0, attname = "history") 
forecast_issue_time <- "2020-02-13T21:32:52-05:00"

nc_close(nc)
```

## Saving to a standardized output format (Option 1)

```{r}
n_depths <- length(depth)
state_names <- c("temperature", "oxygen")
states <- list(temperature, oxygen)
n_states <- length(state_names)
df_combined <- list()

for(k in 1:n_states){
for(i in 1:n_depths){
    df <- as_tibble(states[[k]][, ,i])
    names(df) <- as.character(seq(1, ncol(states[[k]][, ,i])))
    df <- cbind(time, df, data_assimilation)
    df <- df %>% 
      pivot_longer(cols = -c(time,data_assimilation), 
                   names_to = "ensemble", 
                   values_to = state_names[k]) %>% 
      mutate(ensemble = as.integer(ensemble)) %>% 
      mutate(depth = depth[i])
    if(i == 1){
    running_df <- df
    }else{
      running_df <- rbind(running_df, df)
    }
}
      df_combined[[k]] <- running_df
}

df_combined <- right_join(df_combined[[1]], df_combined[[2]], 
                          by = c("time", "ensemble", "depth", "data_assimilation")) %>% 
    mutate(forecast_issue_time = forecast_issue_time,
         Forecast_id = Forecast_id,
         ForecastProject_id = ForecastProject_id,
         scenario = "oxygen_off") %>% 
  select(time, depth, scenario, ensemble, temperature, 
         oxygen, forecast_issue_time, 
         data_assimilation, Forecast_id, ForecastProject_id) 

df_combined_scenario_2 <- df_combined

```

Second, build a data frame for all variables without a depth dimension (parameters).

```{r}
df <- as_tibble(zone1temp[ , ])
names(df) <- as.character(seq(1, ncol(zone1temp)))
df <- cbind(time, data_assimilation, df)
df_combined_p1 <- df %>% 
      pivot_longer(cols = -c("time", "data_assimilation"), 
                   names_to = "ensemble", 
                   values_to = "zone1temp") %>% 
      mutate(ensemble = as.integer(ensemble),
             scenario = "oxygen_off") 

df <- as_tibble(zone2temp[ , ])
names(df) <- as.character(seq(1, ncol(zone2temp)))
df <- cbind(time, data_assimilation, df)
df_combined_p2 <- df %>% 
      pivot_longer(cols = -c("time", "data_assimilation"), 
                   names_to = "ensemble", 
                   values_to = "zone2temp") %>% 
      mutate(ensemble = as.integer(ensemble),
             scenario = "oxygen_off") 

df <- as_tibble(sw_factor[ , ])
names(df) <- as.character(seq(1, ncol(sw_factor)))
df <- cbind(time, data_assimilation, df)
df_combined_p3 <- df %>% 
      pivot_longer(cols = -c("time", "data_assimilation"), 
                   names_to = "ensemble", 
                   values_to = "sw_factor") %>% 
      mutate(ensemble = as.integer(ensemble),
             scenario = "oxygen_off") 

df <- as_tibble(inflow_factor[ , ])
names(df) <- as.character(seq(1, ncol(inflow_factor)))
df <- cbind(time, data_assimilation, df)
df_combined_p4 <- df %>% 
      pivot_longer(cols = -c("time", "data_assimilation"), 
                   names_to = "ensemble", 
                   values_to = "inflow_factor") %>% 
      mutate(ensemble = as.integer(ensemble),
             scenario = "oxygen_off") 

df_parameters_2 <- right_join(df_combined_p1, 
                              df_combined_p2, 
                              by = c("time",
                                     "data_assimilation",
                                     "ensemble",
                                     "scenario")) %>% 
                  right_join(df_combined_p3, 
                              by = c("time",
                                     "data_assimilation",
                                     "ensemble",
                                     "scenario")) %>%
                  right_join(df_combined_p4, 
                              by = c("time",
                                     "data_assimilation",
                                     "ensemble",
                                     "scenario")) %>% 
  mutate(forecast_issue_time = forecast_issue_time,
         Forecast_id = Forecast_id,
         ForecastProject_id = ForecastProject_id) %>% 
  select(time, scenario, ensemble, forecast_issue_time, 
         data_assimilation, Forecast_id, ForecastProject_id, 
         zone1temp, zone2temp, sw_factor, inflow_factor) 
```

## Saving to a standardized output format (Option 1)

Create the 1D and 0D tables

```{r}

df_combined_0D <- rbind(df_parameters_1, df_parameters_2)
df_combined_1D <- rbind(df_combined_scenario_1, df_combined_scenario_2)
```

Write tables to CSV.

```{r}
write.csv(df_combined_1D, "flare-forecast-ensemble-multi-variable-1D.csv", row.names = FALSE)
write.csv(df_combined_0D, "flare-forecast-ensemble-parameters-0D.csv", row.names = FALSE)
```

## Standardized Metadata

Let's document the metadata of the data table itself. It may well be that we decide an Ecological Forecast has to have specific columns like the ones described above, which would thus correspond to a partially pre-defined attributes table (e.g. the units would probably still be allowed to vary, but format would be the same.)  

Note one weakness of this format is that it assumes all data in a column have the same units.  This common assumption might be violoated by transformations to "long" form data, where you have columns like "variable", "value", and "units".  (The long form may be useful, but it exposes much less information in the metadata layer -- e.g. we no longer know what's actually being measured without looking at the data file itself).  

```{r}
#attributes <- tibble::tribble(
#  ~attributeName, ~attributeDefinition, ~unit, ~formatString, ~numberType,
#  "time",          "time",                       "year",     "YYYY-MM-DD%THH:MM_SS%Z", "numberType",
#  "depth",         "depth in reservior",         "meter",   NA,          "real",
#  "ensemble",      "index of ensemble member",   "dimensionless",    NA,         "integer",
#  "scenario",      "forecast scenario",   "dimensionless",    NA, NA,  NA, "scenario name",
#  
#  ##### EDIT STARTING HERE
#  "temperature",     "water temperature", "celsius", NA,  "real", NA,
#  "oxygen",     "oxygen concentration", "numberPerMeterSquared", NA,  "real", NA,
#  #### EDIT ENDING HERE
#  
#  "forecast_issue_time",     "time that forecast was created", "dimensionless", "YYYY-MM-DD",  "numberType",
#  "data_assimilation",     "Flag whether time step included data assimilation", "dimensionless", NA, "integer",
#  "Forecast_id",     "ID for specific forecast cycle", "dimensionless", NA,  "character",
#  "ForecastProject_id",     "ID for forecasting project", "dimensionless", NA,  "character",
#  )

attributes <- tibble::tribble(
  ~attributeName, ~attributeDefinition, ~unit, ~formatString, ~numberType, ~definition,
  "time",          "time",                       "year",     "YYYY-MM-DD", "numberType", NA,
  "depth",         "depth in reservior",         "meter",   NA,          "real", NA,
  "ensemble",      "index of ensemble member",   "dimensionless",    NA,         "integer", NA,
   "scenario",      "forecast scenario",    NA, NA,  NA, "scenario name",
  "temperature",     "water temperature", "celsius", NA,  "real", NA,
  "oxygen",     "oxygen concentration", "numberPerMeterSquared", NA,  "real", NA,
  "forecast_issue_time",     "time that forecast was created", NA, "YYYY-MM-DD",  NA, NA,
  "data_assimilation",     "Flag whether time step included data assimilation", "dimensionless", NA, "integer", NA,
  "Forecast_id",     "ID for specific forecast cycle", NA, NA,  NA, "forecast id",
  "ForecastProject_id",     "ID for forecasting project", NA, NA,  NA, "project id"
)


attrList <- set_attributes(attributes, 
                           col_classes = c("Date", "numeric", "character", "numeric", 
                                           #EDIT STARTING HERE 
                                           "numeric","numeric",
                                           #EDIT ENDING HERE
                                           "Date",
                                           "numeric", "character", "character"))
physical <- set_physical("flare-forecast-ensemble-multi-variable-1D.csv")

dataTable <- eml$dataTable(
                 entityName = "flare-forecast-ensemble-multi-variable-1D.csv",
                 entityDescription = "Falling Creek Reservior Forecast",
                 physical = physical,
                 attributeList = attrList)
```


There's a lot more optional terminology that could be exploited here -- for instance, the specification lets us define different missing value codes (and explanations) for each column, and allows us to indicate `precision`, `minimum` and `maximum`.  

Note that `physical` type can document almost any formats as well, including NetCDF etc.  A NetCDF file would still document the variables measured in much the same way regardless of the underlying representation.  Note that 

Now that we've documented the actual data.frame itself, we can add additional metadata to the record describing our forecast, which is essential for citing, discovering, and interpreting the result.  We start with some authorship information. 

```{r}

me <- list(individualName = list(givenName = "Quinn", 
                                 surName = "Thomas"),
           electronicMailAddress = "rqthomas@vt.edu",
           id = "https://orcid.org/0000-0003-1282-7825")


```

Set Taxonomic, Temporal, and Geographic Coverage.

```{r}
coverage <- 
  set_coverage(begin = as_datetime(time[1]), 
               end = as_datetime(tail((time)[1])),
               geographicDescription = "Falling Creek Reservior",
               west = -79.9, east = -79.9, 
               north = 37.27, south = 37.27)
```

Set key words.  We will need to develop a EFI controlled vocabulary

```{r}
keywordSet <- list(
    list(
        keywordThesaurus = "EFI controlled vocabulary",
        keyword = list("forecast",
                    "ecosystem",
                    "timeseries")
    ))
```


Our dataset needs an abstract describing what this is all about.  Also, a methods section is not required but it's probably a good idea.  Here we import a methods section that was written in Markdown.

**Forecast timestep**: 1 day

**Forecast time horizon** 16 days

**Data assimilation**

  * Data Assimilation used: Yes
  * If, DA used - type of method: EnKF
  * If, DA used - Number of parameters calibrated: 3
  * If, DA used - Sources of training data (DOI, GitHub): https://github.com/CareyLabVT/SCCData/tree/carina-data
  
**Model Description**

  * Type of model (Empirical, process-based, machine learning): Process-based
  * Model name: General Lake Model-AED V3
  * Location of repository with model code: 
    	* GLM: https://github.com/AquaticEcoDynamics/GLM
	    * AED: https://github.com/AquaticEcoDynamics/libaed2
  * Model citation: Hipsey et al. 2019 GMD
  * Total number of model process parameters: Hard to count
  
**Model Covariates**

  * Type (i.e., meteorology): meteorology, Stream Inflow
  * Source (i.e., NOAA GEFS): NOAA GEFS 16-day, https://github.com/CareyLabVT/SCCData/tree/diana-data
  
**Uncertainty (No, Derived from data, Propagates, Assimilates)**

  * Initial conditions: Assimilates
  * Parameter: Assimilates
  * Parameter Random Effects: No
  * Process (within model): Propagates, Assimilates
  * Multi-model: No
  * Driver: Derived from data
  * Scenario: Yes - oxygen system on and off
  * Method for propagating uncertainty (Analytic, ensemble numeric): ensemble numeric
  * If Analytic, specific method
  * If ensemble numeric, number of ensembles: 210
 
```{r}
abstract <- list(markdown = paste(readLines("abstract.md"), collapse = "\n"))
methods <- list(methodStep = list(description = list(markdown = paste(readLines("methods.md"), collapse = "\n"))))
```


```{r}
dataset = eml$dataset(
               title = "FLARE forecast",
               creator = me,
               contact = list(references="https://orcid.org/0000-0003-1282-7825"),
               pubDate = forecast_issue_time,
               intellectualRights = "http://www.lternet.edu/data/netpolicy.html.",
               abstract =  "An illustration of how we might use EML metadata to describe an ecological forecast",
               dataTable = dataTable,
               keywordSet = keywordSet,
               coverage = coverage,
               methods = methods
               )
```

All we need now is to add a unique identifier for the project and we are good to go!  This could be a DOI or merely an identifier we create, e.g. a UUID.

```{r}
my_eml <- eml$eml(dataset = dataset,
           packageId = ForecastProject_id,  
           system = "uuid"
           )
```

Once we have finished building our EML metadata, we can confirm it is valid.  This will catch any missing elements.  (Recall that what is 'required' depends on what you include -- for example, you don't have to document a `dataTable` at all, but if you do, you have to document the "physical" file format it is in  (e.g. `csv`) and the attributes and units it uses!)

```{r}
eml_validate(my_eml)
```

We are now ready to write out a valid EML document: 

```{r}
write_eml(my_eml, "forecast-flare-eml.xml")
```

At this point, we could easily upload this metadata along with the data itself to DataONE via the API (or `dataone` R package.)

We can also generate a JSON-LD version of EML:

```{r}
emld::as_json(as_emld("forecast-flare-eml.xml"), file = "forecast-flare-eml.json")
```

