---
title: "Example metadata for a logistic population growth forecast"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Logistic Growth Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, message=FALSE}
library(EML)
library(uuid)
library(ncdf4)
library(emld)
library(lubridate)
library(tibble)
library(dplyr)
library(tidyr)
emld::eml_version("eml-2.2.0")
set.seed(42)
```

A simple, example forecast of population Growth of two interacting species

========================================================

First, set the forecast identifiers

ForecastProject_id represents the launch of an automated, iterative forecast.  It is created
each time a human modifies the forecast code.  It can be a DOI because this is the level that
we envision citations occuring.

Forecast_id represents each forecast cycle within a ForecastProject_id

For example, if you have a forecast code base on GitHub and launch a forecast from that code that
runs daily for 365 days, then there will be one ForecastProject_id and 365 Forecast_ids. A paper 
analyzing the forecasts would cite the ForecastProject_id.


``` {r}
forecast_issue_time <- as.Date("2001-03-04")
#Forecast_id <- uuid::UUIDgenerate() #ID that applies to the specific forecast
Forecast_id <- "20010304T060000"     # ISO datetime should make a valid Forecast_id
ForecastProject_id <- 30405043 #Some ID that applies to a set of forecasts
```

## Generating the forecast

Multi-species growth for multiple depths with process uncertainty

```{r}
NT <- 30
n_ensembles <- 10
n_depths <- 3
depths <- c(1, 3, 5)

r <- c(1, 3)
K <-  c(10, 20)
alpha <-  c(0.2, 0.3)
n0 <-  0.5

n <- array(NA,dim = c(2, NT, n_depths, n_ensembles))
n[,1,,] <- n0

process_sd <- 0.01

data_assimilation <- rep(0, NT)  

for(t in 2:NT){
  data_assimilation[t] <- 0
  for(depth in 1:n_depths){
    for(ens in 1:n_ensembles){
      n[1, t, depth, ens] <-  n[1, t-1, depth, ens] + 
          r[1]*n[1, t-1,depth,ens]*(1-((n[1, t-1, depth, ens] + 
          alpha[1]*n[2, t-1, depth, ens])/K[1])) + rnorm(1, 0, process_sd)
      n[2, t, depth, ens] <-  n[2, t-1, depth, ens] + 
          r[2]*n[2, t-1, depth, ens]*(1-((n[2, t-1, depth, ens] + 
          alpha[2]*n[1, t-1, depth, ens])/K[2])) + rnorm(1, 0, process_sd)
    }
  }
}
```

## Saving to a standardized output format

# Standard Option 1: netCDF

Convert to a netcdf format 

```{r}
library(ncdf4)

ncfname <- system.file("extdata", "logistic-forecast-ensemble-multi-variable-space-long.nc", package="EFIstandards")

time <- as.Date(as.character(2000 + 1:NT), format = "%Y")

data_assimilation <- rep(0, length(time))

#Set dimensions
ens <- as.integer(seq(1,n_ensembles,1))
depths <- as.integer(c(1,2,3))
timestep <- as.integer(seq(1, NT, 1))

ensdim <- ncdim_def("ens", 
                    units = "",
                    vals = ens, 
                    longname = 'ensemble member') 
depthdim <- ncdim_def("depth", 
                      units = "meters",
                      vals = depths, 
                      longname = 'Depth from surface') 
timedim <- ncdim_def("timestep", 
                     units = '1 day', 
                     longname = 'timestep',
                     vals = timestep)

dimnchar   <- ncdim_def("nchar",   "", 
                        1:nchar(as.character(time[1])), 
                        create_dimvar=FALSE)

#Define variables
fillvalue <- 1e32

def_list <- list()
def_list[[1]] <- ncvar_def(name = "time",
                           units = "datetime",
                           dim = list(dimnchar, timedim),
                           longname = "time",
                           prec="char")
def_list[[2]] <- ncvar_def(name =  "species_1",
                           units = "number of individuals",
                           dim = list(timedim, depthdim, ensdim),
                           missval = fillvalue,
                           longname = 'temperature_mean',
                           prec="single")
def_list[[3]] <- ncvar_def(name =  "species_2",
                           units = "number of individuals",
                           dim = list(timedim, depthdim, ensdim),
                           missval = fillvalue,
                           longname = 'temperature_mean',
                           prec="single")
def_list[[4]] <- ncvar_def(name =  "data_assimilation",
                           units = "logical",
                           dim = list(timedim),
                           missval = fillvalue,
                           longname = '1 = data assimilation used in timestep',
                           prec="single")

ncout <- nc_create(ncfname,def_list,force_v4=T)

ncvar_put(ncout,def_list[[1]] , time)
ncvar_put(ncout,def_list[[2]] , n[1, , , ])
ncvar_put(ncout,def_list[[3]] , n[2, , , ])
ncvar_put(ncout,def_list[[4]] , data_assimilation)

#Global file metadata
ncatt_put(ncout,0,"ForecastProject_id", as.character(ForecastProject_id), 
          prec =  "text")
ncatt_put(ncout,0,"Forecast_id",as.character(Forecast_id), 
          prec =  "text")
ncatt_put(ncout,0,"forecast_issue_time",as.character(forecast_issue_time), 
          prec =  "text")
nc_close(ncout)
```

# Standard Option 2: ensemble CSV

Convert to a flat file format (CSV) with one column for each variable and all
ensemble members saved

```{r}
time <- as.Date(as.character(2000 + 1:NT), format = "%Y")
state_names <- c("species_1", "species_2")
n_states <- length(state_names)
states <- list(n[1, , , ], n[2, , ,])

#No data assimilation was used for any of forecast results archived

df_combined <- list()

for(k in 1:n_states){
  for(i in 1:n_depths){
    df <- as_tibble(states[[k]][, ,i])
    names(df) <- as.character(seq(1, ncol(states[[k]][, ,i])))
    df <- cbind(time, df, data_assimilation)
    df <- df %>% 
      pivot_longer(cols = -c(time,data_assimilation), 
                   names_to = "ensemble", 
                   values_to = state_names[k]) %>% 
      mutate(ensemble = as.integer(ensemble)) %>% 
      mutate(depth = depths[i])
    if(i == 1){
      running_df <- df
    }else{
      running_df <- rbind(running_df, df)
    }
  }
  df_combined[[k]] <- running_df
}

df_combined <- right_join(df_combined[[1]], df_combined[[2]], 
                          by = c("time", "ensemble", "depth", "data_assimilation")) %>% 
  mutate(forecast_issue_time = forecast_issue_time,
         Forecast_id = Forecast_id,
         ForecastProject_id = ForecastProject_id) %>% 
  select(time, depth, ensemble, state_names[1], 
         state_names[2], forecast_issue_time, 
         data_assimilation, ForecastProject_id, Forecast_id) 

df_combined

write.csv(df_combined, 
          file = "logistic-forecast-ensemble-multi-variable-multi-depth.csv")
```

# Standard Option 3: summary CSV

Convert to a flat file format (CSV) with forecast distribution summaries saved

```{r}
df_species_1 <- df_combined %>% 
  select(-species_2) %>% 
  group_by(time, depth, forecast_issue_time, data_assimilation, 
           ForecastProject_id, Forecast_id) %>% 
  summarize(mean = mean(species_1),
            Conf_interv_02.5 = quantile(species_1, 0.025),
            Conf_interv_97.5 = quantile(species_1, 0.975)) %>% 
  pivot_longer(cols = c("mean","Conf_interv_02.5","Conf_interv_97.5"),
               names_to = "Statistic",
               values_to = "species_1")

df_species_2 <- df_combined %>% 
  select(-species_1) %>% 
  group_by(time, depth, forecast_issue_time, data_assimilation, 
           ForecastProject_id, Forecast_id) %>% 
  summarize(mean = mean(species_2),
            Conf_interv_02.5 = quantile(species_2, 0.025),
            Conf_interv_97.5 = quantile(species_2, 0.975)) %>% 
  pivot_longer(cols = c("mean","Conf_interv_02.5","Conf_interv_97.5"),
               names_to = "Statistic",
               values_to = "species_2")

 df_summary <- right_join(df_species_1, df_species_2)
 
 df_summary
 
write.csv(df_summary, 
          file = "logistic-forecast-summary-multi-variable-multi-depth.csv")

```

## Standardized Metadata

Let's document the metadata of the data table itself. It may well be that we 
decide an Ecological Forecast has to have specific columns like the ones 
described above, which would thus correspond to a partially pre-defined 
attributes table (e.g. the units would probably still be allowed to vary, but 
format would be the same.)  

Note one weakness of this format is that it assumes all data in a column have 
the same units.  This common assumption might be violoated by transformations 
to "long" form data, where you have columns like "variable", "value", and 
"units".  (The long form may be useful, but it exposes much less information in 
the metadata layer -- e.g. we no longer know what's actually being measured 
without looking at the data file itself).  

```{r}
attributes <- tibble::tribble(
  ~attributeName, ~attributeDefinition, ~unit, ~formatString, ~numberType, ~definition,
  "time",          "time",                       "year",     "YYYY-MM-DD", "numberType", NA,
  "depth",         "depth in reservior",         "meter",   NA,          "real", NA,
  "ensemble",      "index of ensemble member",   "dimensionless",    NA,         "integer", NA,
  "species_1",     "Population size of species 1", "numberPerMeterSquared", NA,  "real", NA,
  "species_2",     "Population size of species 2", "numberPerMeterSquared", NA,  "real", NA,
  "forecast_issue_time",     "time that forecast was created", NA, "YYYY-MM-DD",  NA, NA,
  "data_assimilation",     "Flag whether time step included data assimilation", "dimensionless", NA, "integer", NA,
  "Forecast_id",     "ID for specific forecast cycle", NA, NA,  NA, "forecast id",
  "ForecastProject_id",     "ID for forecasting project", NA, NA,  NA, "project id"
)
attrList <- set_attributes(attributes, 
                           col_classes = c("Date", "numeric", "numeric", 
                                           "numeric","numeric", "Date",
                                           "numeric", "character", "character"))
physical <- set_physical("logistic-forecast-ensemble-multi-variable-multi-depth.csv")

dataTable <- eml$dataTable(
                 entityName = "logistic-forecast-ensemble-multi-variable-multi-depth.csv",
                 entityDescription = "Forecast of population size using a depth specific model",
                 physical = physical,
                 attributeList = attrList)
```


There's a lot more optional terminology that could be exploited here -- for 
instance, the specification lets us define different missing value codes (and 
explanations) for each column, and allows us to indicate `precision`, `minimum`
and `maximum`.  

Note that `physical` type can document almost any formats as well, including 
NetCDF etc.  A NetCDF file would still document the variables measured in much 
the same way regardless of the underlying representation.  Note that 

Now that we've documented the actual data.frame itself, we can add additional 
metadata to the record describing our forecast, which is essential for citing, 
discovering, and interpreting the result.  We start with some authorship 
information. 

```{r}

me <- list(individualName = list(givenName = "Quinn", 
                                 surName = "Thomas"),
           electronicMailAddress = "rqthomas@vt.edu",
           id = "https://orcid.org/0000-0003-1282-7825")


```


Set Taxonomic, Temporal, and Geographic Coverage.  (Look, apparently we're 
modeling population densities of *Sarracenia purpurea* in Harvard Forest 
starting in about 2012!)

```{r}
coverage <- 
  set_coverage(begin = '2012-06-01', 
               end = '2013-12-31',
               sci_names = "Sarracenia purpurea",
               geographicDescription = "Harvard Forest Greenhouse, Tom Swamp Tract (Harvard Forest)",
               west = -122.44, east = -117.15, 
               north = 37.38, south = 30.00,
               altitudeMin = 160, altitudeMaximum = 330,
               altitudeUnits = "meter")
```

Set key words.  We will need to develop a EFI controlled vocabulary

```{r}
keywordSet <- list(
    list(
        keywordThesaurus = "EFI controlled vocabulary",
        keyword = list("forecast",
                    "population",
                    "timeseries")
    ))
```


Our dataset needs an abstract describing what this is all about.  Also, a 
methods section is not required but it's probably a good idea.

We envision having additional required metadata sections.  However, we need
to figure out the formatting of the Markdown file so that it is represented 
cleanly in the EML.  

```{r}
additionalMetadata <- eml$additionalMetadata(
  #  describes="forecast",  ## not sure how to find the correct ID for this to be valid
  metadata = list(
    forecast = list(
      timestep = "1 year", ## should be udunits parsable; already in coverage -> temporalCoverage?
      forecast_horizon = "30 years",
      initial_conditions = list(
        # Possible values: no, contains, data_driven, propagates, assimilates
        uncertainty = "contains",
        # Number of parameters / dimensionality
        complexity = 2
      ),
      parameters = list(
        uncertainty = "contains",
        complexity = 3
      ),
      random_effects = list(
        uncertainty = "no"
      ),
      process_error = list(
        uncertainty = "propagates",
        propagation = list(
          type = "ensemble", # ensemble vs analytic
          size = 10          # required if ensemble
        ),
        complexity = 2,
        covariance = FALSE
      ),
      drivers = list(
        uncertainty = "no"
      )
      # assimilation_method ## required if any uncertainty = assimilates
    ) # forecast
  ) # metadata
) # eml$additionalMetadata

```


**Model Description**

  * Type of model (Empirical, process-based, machine learning): Process-based
  * Model name: discrete Lotka–Volterra model
  * Location of repository with model code: https://github.com/somewhere or https://doi.org/10.xxx
  * Model citation: N/A
  * Total number of model process parameters: 3
  
**Model Covariates**

  * Type (i.e., meteorology): N/A
  * Source (i.e., NOAA GEFS): N/A
  
```{r}
abstract_text <- system.file("extdata", "abstract.md", package="EFIstandards", mustWork = TRUE)
methods_text <- system.file("extdata", "methods.md",  package="EFIstandards", mustWork = TRUE)

abstract <- list(markdown = paste(readLines(abstract_text), collapse = "\n"))
methods <- list(id="forecast", 
                methodStep = list(description = 
                                    list(markdown = paste(readLines(methods_text), 
                                                          collapse = "\n"))))  ## to be dropped
```


```{r}
dataset = eml$dataset(
               title = "A very silly logistic forecast",
               creator = me,
               contact = list(references="https://orcid.org/0000-0003-1282-7825"),
               pubDate = forecast_issue_time,
               intellectualRights = "http://www.lternet.edu/data/netpolicy.html.",
               abstract =  "An illustration of how we might use EML metadata to describe an ecological forecast",
               dataTable = dataTable,
               keywordSet = keywordSet,
               coverage = coverage,
               methods = methods
               )
```


All we need now is to add a unique identifier for the project and we are good to
go!  This could be a DOI or merely an identifier we create, e.g. a UUID.

```{r}
my_eml <- eml$eml(dataset = dataset,
           additionalMetadata = additionalMetadata,
           packageId = Forecast_id,  #Is this the ForecastProject_ID?
           #What about the Forecast_ID
           system = "uuid"
           )
```


Once we have finished building our EML metadata, we can confirm it is valid.  
This will catch any missing elements.  (Recall that what is 'required' depends 
on what you include -- for example, you don't have to document a `dataTable` at 
all, but if you do, you have to document the "physical" file format it is in  
(e.g. `csv`) and the attributes and units it uses!)

```{r}
eml_validate(my_eml)
```

We are now ready to write out a valid EML document: 

```{r}
write_eml(my_eml, "forecast-eml.xml")
```

At this point, we could easily upload this metadata along with the data itself 
to DataONE via the API (or `dataone` R package.)

We can also generate a JSON-LD version of EML:

```{r}
emld::as_json(as_emld("forecast-eml.xml"), file = "forecast-eml.json")
```





```{r include=FALSE}
## Cleanup
lapply(list.files(pattern = "[.]csv"), unlink)
lapply(list.files(pattern = "[.]json"), unlink)
lapply(list.files(pattern = "[.]xml"), unlink)

```
